Complete draft

### 2019-06-10 doc/point.tex
more edits before sending to Chao

There is sufficient information in the distribution of peculiar velocities and overdensities of galaxies to test general relativity on the largest scale. Suppose we have a region that encloses $N_{gal}$ number of galaxies in the universe. We can divide the region into $N_{cell}\gg 1$ number of equal volume cells whose size can be made so small that either one or no galaxy is contained in one cell. This is encoded in a vector $\textbf{N}$ of size $N_{cell}$ whose entries are either 0 or 1. For any 2 cells, we can compute their correlation, a function of the distance $r$ between the cells only. In our application, we consider $\xi_{\delta\delta}$ (the overdensity-overdensity correlation), $\xi_{\delta v}$ (overdensity-velocity correlation), $\xi_{v\delta}$ (velocity-overdensity correlation), and $\xi_{vv}$ (velocity-velocity correlation).\par

%The data  is $\textbf{x}= [N_1-\bar{N}, N_2-\bar{N}, \hdots, N_{N_{cell}}-\bar{N}, \Delta v_1, \hdots, \Delta v_{N_{gal}}]^T$, where $\Delta v$ is the deviation of the measured velocity from the expected, 
The data themselves are denoted as $\textbf{y} = [N_1, N_2, \hdots, N_{N_{cell}}, v_1, \hdots, v_{N_{gal}}]$
and are modeled as $\textbf{y}_0 =[\bar{N}, \bar{N}, \hdots, \bar{N}, v^0_{1}, \hdots, v^0_{N_{gal}}]$.  For now we assume that the model is not fixed but depends
on linear parameters $p_\alpha$ such that
 $\textbf{y}_0 = p_\alpha \textbf{x}_\alpha$.
Their difference is  denoted as $\Delta \textbf{y}= \textbf{y} - \textbf{y}_0 $ and $\Delta v= v-v^0$.  (Although the note is written in terms of velocities
in practice we expect to work with magnitudes as the are observed.)

We postulate that deviations from expected variance $\bar{N}$ and expected peculiar velocity $\bar{v}$ are Gaussian. Then the above correlation matrices can be used to construct a block matrix $\textbf{C}$, parameterized by $A$ and $\bm{\sigma}_v$, a vector of shot noise for peculiar velocity measurement:
\begin{align}
\textbf{C} &\equiv \textbf{C}_0 + \textbf{S}\\
%\textbf{C} &\equiv \textbf{C}_0 + A\textbf{S} \\ 
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+
\begin{pmatrix}
\bar{N}^2\xi_{gg} & \bar{N}\xi_{g v} \\
\bar{N}\xi_{v g} & \xi_{vv}
\end{pmatrix},
\end{align}
where $\xi_{gg}$, $\xi_{vv}$, and $\bar{N}\xi_{g v}$ are the galaxy overdensity and peculiar velocity correlation functions and their cross-correlation
respectively.

The reason we have $\bar{N}$ as the variance for overdensity is that our small galaxy cells are best modeled by Poisson statistics, in which the mean equals the variance. In addition, note that the dimension of $C$ is $N_{cell}+N_{gal}$ by $N_{cell}+N_{gal}$. For the peculiar velocity covariance, we assign $\sigma^2=\infty$ to cells that do not contain any galaxy, and one can show that removing rows and columns that contain those cells has no effect but a constant offset on what we will compute later.

The science parameters $\theta$ enter through these correlation functions, and may include $\theta \in \{fD, bD, \gamma, \Omega_{M_0},\ldots\}$ and others
that describe the matter power spectrum.
For example, the parameter $fD$ enters in through the relations $\xi_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy count overdensity
power spectrum $\xi_{g g }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $\xi_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu\equiv \cos{(\hat{k} \cdot \hat{r})}$ where $\hat{r}$ is the direction of
the line of sight.  
Assuming linear dependence of the parameters,
\begin{align}
\textbf{C} 
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+ \sum_\theta \theta
\begin{pmatrix}
\bar{N}^2\xi_{gg,\theta} & \bar{N}\xi_{g v,\theta} \\
\bar{N}\xi_{v g,\theta} & \xi_{vv,\theta}
\end{pmatrix}\\
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+ \sum_\alpha \theta_\alpha \textbf{S}_\alpha .
\end{align}

%\left(
%\begin{array}{cc}
%\bar{N}\textbf{I} & 0 \nonumber \\
%0 & \bm{\sigma}_v^2\textbf{I}
%\end{array}
%\right)
%+
%A
%\left(
%\begin{array}{cc}
%\bar{N}^2\xi_{\delta\delta} & \bar{N}\xi_{\delta v} \\
%\bar{N}\xi_{\delta v} & \xi_{vv}
%\end{array}
%\right).
%\end{align}\par

The Gaussian likelihood as a function of $\theta$
\[
\mathcal{L}(\theta) = \frac{1}{\sqrt{\left(2\pi\right)^{N_{cell}+N_{gal}} \det \textbf{C}}}\exp\left\{-\frac{1}{2}\Delta \textbf{y}^T\textbf{C}^{-1}\Delta \textbf{y}\right\}.
\]\par
To make the expression cleaner, we take the log of the likelihood and focus on the $A$-dependent part:
\begin{equation}
-2\ln\mathcal{L}(\theta) = \ln\det\textbf{C} + \Delta \textbf{y}^T\textbf{C}^{-1}\Delta \textbf{y} + \text{const}.
\label{log:eqn}
\end{equation}

The estimator and Fisher matrix for this Gaussian likelihood are
\begin{align}
\hat{\theta}_\alpha & = \frac{1}{2} F^{(\theta \theta)\,-1}_{\alpha \beta} [(\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y}) - \tr(\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{C}_0)]\\
\hat{p}_\alpha  &=  F^{(pp)\, -1}_{\alpha \beta}  (\textbf{y}^T  \textbf{C}^{-1} x_\beta)
\end{align}
where  
\begin{align}
F^{(\theta\theta)}_{\alpha\beta}& = \frac{1}{2} \tr(\textbf{C}^{-1} \textbf{S}_\alpha \textbf{C}^{-1}  \textbf{S}_\beta)\\
F^{(pp)}_{\alpha\beta} & =x^T_\alpha \textbf{C}^{-1} x_\beta\\
F^{(\theta p)}_{\alpha\beta} & = 0.
\end{align}

We focus on the parameters $\theta$ related to the covariance matrix .
The term
\begin{align}
\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y} & \approx \Delta \textbf{y}^T \textbf{C}_0^{-1} \textbf{S}_\beta \textbf{C}_0^{-1} \Delta \textbf{y}\\
&=(\textbf{N}-\bar{N}) \xi_{\delta\delta,\theta_\beta}  (\textbf{N}-\bar{N})  + 2  (\textbf{N}-\bar{N}) \xi_{\delta v,\theta_\beta}\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2}+ 
\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2} \xi_{vv,\theta_\beta}\frac{\mathbf{\Delta v}}{\bm{\sigma}_v^2}.
\end{align}
Noting that $N_i=1$ when the cell $i$ contains a galaxy and zero otherwise and that only cells with a peculiar velocity measurement contribute ($\sigma_v =\infty$
for cells without such a measurement), and that in the limit of infinitely small cells, $\bar{N}\sum_{i,j\in N_{cell}} = \int n dV$, where $n$ is the galaxy number density,
\begin{align*}
\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y} & \approx \sum_{i,j\in N_{gal}} \xi_{\delta\delta,\theta_\beta,ij} -2 \bar{N} \sum_{i\in N_{gal},j\in N_{cell}} \xi_{\delta\delta,\theta_\beta,ij}\\
& +\bar{N}^2  \sum_{i,j\in N_{cell}}  \xi_{\delta\delta,\theta_\beta,ij} + 2 \sum_{i,j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta,ij} \\
& -2\bar{N}\sum_{i\in N_{cell}, j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta,ij}  + \sum_{i,j\in N_{gal}} \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,\theta_\beta,ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\\
& = \sum_{i,j\in N_{gal}} \left( \xi_{\delta\delta,\theta_\beta}(x_i,x_j) + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta}(x_i,x_j) + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,\theta_\beta}(x_i,x_j) \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
&-2  \sum_{i\in N_{gal}}\int n(x_i) \left(\xi_{\delta\delta,\theta_\beta}(x_i,y) + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{v \delta,\theta_\beta}(x_i,y) \right) d^3y\\
&  + \int n(x) n(y) \xi_{\delta\delta,\theta_\beta}(x,y)d^3xd^3y.
\end{align*}

The trace term depends on only the zero-lag correlation
\begin{align}
 \tr(\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{C}_0) & \approx  \tr(\textbf{C}_0^{-1} \textbf{S}_\beta) \\
 & = \bar{N} \tr(\xi_{\delta\delta,\beta}) + \tr(\frac{1}{\sigma^2_v} \xi_{vv,\beta})\\
 & = \int n(x) \xi_{\delta \delta,\beta}(x,x) d^3x +  \sum_{i\in N_{gal}} \frac{1}{\sigma^2_{v,i}} \xi_{vv,\beta}(x_i,x_i).
\end{align}

The term in the Fisher matrix, to the leading term in $\bar{N}$ is
\begin{align}
 \tr(\textbf{C}^{-1} \textbf{S}_\alpha \textbf{C}^{-1} \textbf{S}_\beta) & \approx  \tr(\textbf{C}_0^{-1} \textbf{S}_\alpha \textbf{C}_0^{-1} \textbf{S}_\beta)  \\
 & = \bar{N}^2 \tr(\xi_{\delta \delta,\alpha}\xi_{\delta \delta,\beta})+
 \bar{N} \tr(\xi_{\delta v,\alpha} \frac{1}{\bm{\sigma}^2_v} \xi_{v\delta,\beta} ) + \tr(\frac{1}{\bm{\sigma}^2_v} \xi_{vv,\alpha}  \frac{1}{\bm{\sigma}^2_v}  \xi_{vv,\beta})\\
 & = \int n(x) n(y) \xi_{\delta \delta,\alpha}(x,y)\xi_{\delta \delta,\beta}(y,x) d^3x d^3y \\
 & + \int n(x) \sum_{i\in N_{gal}} \xi_{\delta v,\alpha}(x,x_i) \frac{1}{\sigma^2_{v,i}} \xi_{v\delta,\beta} (x_i,x) d^3x \\
 & +  \sum_{i,j\in N_{gal}} \frac{1}{\sigma^2_{v,i}} \xi_{vv,\alpha}(x_i,x_j)  \frac{1}{\sigma^2_{v,j}}  \xi_{vv,\beta}(x_j,x_i).
\end{align}

>>>>

%There is sufficient information in the distribution of peculiar velocities and overdensities of galaxies to test general relativity on the largest scale. Suppose we have a region that encloses $N_{gal}$ number of galaxies in the universe. We can divide the region into $N_{cell}\gg 1$ number of equal volume cells whose size can be made so small that either one or no galaxy is contained in one cell. This is encoded in a vector $\textbf{N}$ of size $N_{cell}$ whose entries are either 0 or 1. For any 2 cells, we can compute their correlation, a function of the distance $r$ between the cells only. In our application, we consider $\xi_{\delta\delta}$ (the overdensity-overdensity correlation), $\xi_{\delta v}$ (overdensity-velocity correlation), $\xi_{v\delta}$ (velocity-overdensity correlation), and $\xi_{vv}$ (velocity-velocity correlation).\par

Consider a galaxy count and peculiar velocity survey that covers a volume, which is divided into $N_{cell}$ equal-volume cells.  Each cell will contain some
number of survey galaxies; when the cell volume is large then each cell is likely to contain many galaxies whereas when the cell volume is small each cell is likely
to have either one or no galaxies.  Similarly, when the cell volume is large  its  peculiar velocity is determined combining all the velocity measurements therein,
whereas when the volume is small the cell velocity is likely to be that of the one galaxy therein or there is no measurement, alternatively viewed as an infinite
uncertainty measurement.  We denote $N_{gal}$ as the number of cells that contains at least one galaxy, at the large cell volume limit
$N_{gal}=N_{cell}$ whereas in the limit of infinitely small cell volumes it is
the number of galaxies in the survey.

%The data  is $\textbf{x}= [N_1-\bar{N}, N_2-\bar{N}, \hdots, N_{N_{cell}}-\bar{N}, \Delta v_1, \hdots, \Delta v_{N_{gal}}]^T$, where $\Delta v$ is the deviation of the measured velocity from the expected, 
The data in the cells are denoted as $\textbf{y} = [N_1, N_2, \hdots, N_{N_{cell}}, v_1, \hdots, v_{N_{gal}}]$
and are modeled as $\textbf{y}_0 =[\bar{N}, \bar{N}, \hdots, \bar{N}, v_{0,1}, \hdots, v_{0,N_{gal}}]$.  For now we assume that the model is not fixed but depends
on linear parameters $p_\alpha$ such that
 $\textbf{y}_0 = p_\alpha \textbf{x}_\alpha$.
Their difference is  denoted as $\Delta \textbf{y}= \textbf{y} - \textbf{y}_0 $ and $\Delta v= v-v_0$.  (Although the note is written in terms of velocities
in practice we expect to work with magnitudes.)

We model the observations as being drawn from a Gaussian distribution.
The variance in number counts in a cell is $\bar{N}$ and the vector of peculiar velocity measurement in the cells is denoted $\bm{\sigma}_v$,
  Then the above correlation matrices can be used to construct a block matrix $\textbf{C}$:
\begin{align}
\textbf{C} &\equiv \textbf{C}_0 + \textbf{S}\\
%\textbf{C} &\equiv \textbf{C}_0 + A\textbf{S} \\ 
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+
\begin{pmatrix}
\bar{N}^2\xi_{gg} & \bar{N}\xi_{g v} \\
\bar{N}\xi_{v g} & \xi_{vv}
\end{pmatrix},
\end{align}
where $\xi_{gg}$, $\xi_{vv}$, and $\xi_{g v}$ are the galaxy overdensity and peculiar velocity correlation functions and their cross-correlation
respectively.  
The dimension of $C$ is $N_{cell}+N_{gal}$ by $N_{cell}+N_{gal}$. For the peculiar velocity covariance, we assign $\sigma^2=\infty$ to cells that do not contain any galaxy, and one can show that removing rows and columns that contain those cells has no effect but a constant offset on what we will compute later.

The science parameters $\theta$ enter through the correlation functions, and may include $\theta \in \{fD, bD, \gamma, \Omega_{M_0},\ldots\}$ and others
that describe the matter power spectrum.
For example, the parameter $fD$ enters in through the relations $\xi_{vv}\propto (fD\mu)^2$, the SN~Ia host-galaxy count overdensity
power spectrum $\xi_{g g }\propto (bD + fD\mu^2)^2$, and the galaxy-velocity cross-correlation $\xi_{vg}
\propto  (bD + fD\mu^2)fD$, where $b$ is the galaxy bias and $\mu\equiv \cos{(\hat{k} \cdot \hat{r})}$ where $\hat{r}$ is the direction of
the line of sight.   There are also the parameters $p$ that linearly affect the mean of the Gaussian, a primary example being the combination
$M+5\log{H_0}$ where $M$ is the SN absolute magnitude and $H_0$ is the Hubble constant.

The Gaussian likelihood as a function of $\theta$ is
\[
\mathcal{L}(\theta) = \frac{1}{\sqrt{\left(2\pi\right)^{N_{cell}+N_{gal}} \det \textbf{C}}}\exp\left\{-\frac{1}{2}\Delta \textbf{y}^T\textbf{C}^{-1}\Delta \textbf{y}\right\}.
\]
and
\begin{equation}
-2\ln\mathcal{L}(\theta) = \ln\det\textbf{C} + \Delta \textbf{y}^T\textbf{C}^{-1}\Delta \textbf{y} + \text{const}.
\label{log:eqn}
\end{equation}

Assuming linear dependence of the parameters,
\begin{align}
\textbf{C} 
&=
\left(
\begin{array}{cc}
\bar{N}\textbf{I} & 0 \nonumber \\
0 & \bm{\sigma}_v^2\textbf{I}
\end{array}
\right)
+ \sum_\alpha \theta_\alpha
\begin{pmatrix}
\bar{N}^2\xi_{gg,{\theta_\alpha}} & \bar{N}\xi_{g v,{\theta_\alpha}} \\
\bar{N}\xi_{v g,{\theta_\alpha}} & \xi_{vv,{\theta_\alpha}}
\end{pmatrix}\\
&=
\textbf{C}_0
+ \sum_\alpha \theta_\alpha \textbf{S}_\alpha,
\end{align} 
where the comma represents the partial derivative.

%\begin{comment}
%The inverse can be expanded as
%\begin{align}
%\textbf{C}^{-1}
%&=
%\textbf{C}_0^{-1} \left( 1 +  \sum_\alpha \theta_\alpha \textbf{S}_\alpha \textbf{C}_0^{-1} \right)^{-1}\\
%&=
%\textbf{C}_0^{-1} \sum_{i=0}^{\infty} (-1)^i \left(\sum_\alpha \theta_\alpha \textbf{S}_\alpha \textbf{C}_0^{-1} \right)^i.
%\end{align} 
%
%In our case
%\begin{align}
% \textbf{S}_\alpha \textbf{C}_0^{-1} & = 
% \begin{pmatrix}
% \bar{N}\xi_{gg,\theta_\alpha} & \frac{\bar{N}}{ \bm{\sigma}_v^2} \xi_{gv,\theta_\alpha}\\
% \xi_{vg,\theta_\alpha} & \frac{1}{ \bm{\sigma}_v^2} \xi_{vv,\theta_\alpha}\\
% \end{pmatrix}
% \end{align}
%
%We will eventually take the limit where $\bar{N} \rightarrow 0$.
%Then
%\begin{align}
%\left(\sum_\alpha \theta_\alpha \textbf{S}_\alpha \textbf{C}_0^{-1} \right)^i &  \approx 
%\sum_{\alpha}  \ldots \sum_{\delta}  \theta_{\alpha} \ldots \theta_\delta
%  \begin{pmatrix}
% \frac{\bar{N}}{ \bm{\sigma}_v^{2(i-1)}} \xi_{gv,\theta_\alpha}  \xi_{vg,\theta_\beta} \ldots  \xi_{vg,\theta_\gamma} & \frac{\bar{N}}{ \bm{\sigma}_v^{2i}} \xi_{gv,\theta_\alpha} \xi_{vv,\theta_\beta} \ldots  \xi_{vv,\theta_\gamma} \\
% \frac{1}{ \bm{\sigma}_v^{2(i-1)}} \xi_{vv,\theta_\alpha} \xi_{vg,\theta_\beta}\ldots  \xi_{vg,\theta_\gamma}   & \frac{1}{ \bm{\sigma}_v^{2i}} \xi_{vv,\theta_\alpha} \xi_{vv,\theta_\beta}\ldots  \xi_{vv,\theta_\gamma}\\
% \end{pmatrix}.
%\end{align}
%\end{comment}
%\left(
%\begin{array}{cc}
%\bar{N}\textbf{I} & 0 \nonumber \\
%0 & \bm{\sigma}_v^2\textbf{I}
%\end{array}
%\right)
%+
%A
%\left(
%\begin{array}{cc}
%\bar{N}^2\xi_{\delta\delta} & \bar{N}\xi_{\delta v} \\
%\bar{N}\xi_{\delta v} & \xi_{vv}
%\end{array}
%\right).
%\end{align}\par

The estimator and Fisher matrix for this Gaussian likelihood are well known (and derived in  \S\ref{appendix:sec}).
\begin{align}
\hat{\theta}_\alpha & = \frac{1}{2} F^{(\theta \theta)\,-1}_{\alpha \beta} [(\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y}) - \tr(\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{C}_0)] \nonumber \\
\hat{p}_\alpha  &=  F^{(pp)\, -1}_{\alpha \beta}  (\textbf{y}^T  \textbf{C}^{-1} x_\beta)
\label{estimators:eqn}
\end{align}
where  
\begin{align}
F^{(\theta\theta)}_{\alpha\beta}& = \frac{1}{2} \tr(\textbf{C}^{-1} \textbf{S}_\alpha \textbf{C}^{-1}  \textbf{S}_\beta)\\
F^{(pp)}_{\alpha\beta} & =x^T_\alpha \textbf{C}^{-1} x_\beta\\
F^{(\theta p)}_{\alpha\beta} & = 0.
\end{align}

%Let us consider the limit where there are an infinite number of cells and $\bar{N} \rightarrow 0$ and $\bar{N}\sum_{i,j\in N_{cell}} = n \int dV$, where $n$ is the galaxy number density,
Denoting $\textbf{M}_\beta = \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1}$ and splitting the calculation into quadrants  corresponding to the galaxy and velocity indices,
the complicated terms are
\begin{align*}
\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y}  
 %& = \frac{1}{\bar{N}^2}\int \int n(x_1) n(x_2) (N(x_1)-\bar{N}) M_{gg}(x_1,x_2) (N(x_2)-\bar{N})  d^3x_1  d^3x_2\\
%& +   \frac{2}{\bar{N}}\int \sum_{i=1}^{N} n(x_1) (N(x_1)-\bar{N}) M_{gv}(x_1,x_i) \Delta v_i  d^3x_1 \\
%& + \sum_{i=1}^{N}  \sum_{j=1}^{N}  \Delta v_i M_{vv}(x_i,x_j) \Delta v_j\\
&= \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}} N_i M_{\beta,gg}(x_i,x_j) N_j 
 - 2\bar{N}   \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text gal}}  M_{\beta,gg}(x_i,x_j) N_j \\
& +\bar{N}^2 \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text cell}}  M_{\beta,gg}(x_i,x_j) 
 +   2 \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}}  N_i M_{\beta,gv}(x_i,x_j) \Delta v_j  \\
& -   2 \bar{N}  \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text gal}}  M_{\beta,gv}(x_i,x_j) \Delta v_j 
+ \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}}  \Delta v_i M_{\beta,vv}(x_i,x_j) \Delta v_j ,
\end{align*}
\begin{align*}
\tr (\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{S}_\alpha) & 
=  \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text cell}} M_{\beta,gg}(x_i,x_j)\textbf{S}_{\alpha,gg}(x_j,x_i)  +  2 \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text gal}} M_{\beta,gv}(x_i,x_j)\textbf{S}_{\alpha,vg} (x_j,x_i)\\
& + \sum_{i=1}^{N_{\text gal}} \sum_{j=1}^{N_{\text gal}} M_{\beta,vv}(x_i,x_j) \textbf{S}_{\alpha,vv}(x_j,x_i),
\end{align*}
\begin{align*}
\tr (\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{C}_0) & 
=  \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text cell}} M_{\beta,gg}(x_i,x_j)\textbf{C}_{0,gg}(x_j,x_i)  + \sum_{i=1}^{N_{\text gal}} \sum_{j=1}^{N_{\text gal}} M_{\beta,vv}(x_i,x_j) \textbf{C}_{0,vv}(x_j,x_i)\\
 & 
= \bar{N} \sum_{i=1}^{N_{\text cell}} M_{\beta,gg}(x_i,x_i)  + \sum_{i=1}^{N_{\text gal}}  M_{\beta,vv}(x_i,x_i) \sigma^{2}_{vi} .
\end{align*}

An interesting limit is when there are infinitely small cell sizes.  Then $N_{gal}$ (the number of cells with galaxies) approaches the number
of galaxies, those cells have $N_i=1$,  and $\bar{N}\sum_{i\in N_{cell}} = n \int dV$, where $n$ is the galaxy number density.
The calculation then is then `binless', with sums over sample objects and integrals to account for the volume where there are no objects.
In this limit, a Poisson rather than Gaussian model distribution better describes expectations.

Finding the roots of Eq.~\ref{estimators:eqn} to get the estimators is non-trivial.  In particular the $\hat{\theta}$ appears in $\mathbf{C}^{-1}$, a large matrix that
isn't obviously easy to invert.  Using the renormalization group, \cite{2019PhRvD..99d3538M} reparameterizes the likelihood such that much of the parameter-dependence
gets shifted to other terms leaving a smaller
band-diagonal matrix that remains to be inverted.  Applying his methodology to our problem is an interesting direction to pursue.

For our purposes, having a precise answer is probably unnecessary.
Solving the right hand side of Eq.~\ref{estimators:eqn} to lowest (zeroth) order in $\theta$ gives a useful approximate answer.  Replacing
$\textbf{C}^{-1}$ with the diagonal $\textbf{C}_0^{-1}$ gives
\begin{align*}
\textbf{M}_{\beta,gg} & \approx \xi_{gg,\theta_\beta}\\
\textbf{M}_{\beta,gv} & \approx \xi_{gv,\theta_\beta} \bm{\sigma}^{-2}_v\\
\textbf{M}_{\beta,vv} & \approx \bm{\sigma}^{-2}_v \xi_{vv,\theta_\beta} \bm{\sigma}^{-2}_v .
\end{align*}
so that
\begin{align*}
\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y}  
&\approx \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}} N_i \xi_{gg,\theta_\beta}(x_i,x_j) N_j 
 - 2\bar{N}   \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text gal}}  \xi_{gg,\theta_\beta}(x_i,x_j)N_j \\
& +\bar{N}^2 \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text cell}} \xi_{gg,\theta_\beta}(x_i,x_j)
 +   2 \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}}  N_i  \xi_{gv,\theta_\beta}(x_i,x_j)\frac{\Delta v_j}{\sigma^2_{v,j}} \\
& -   2 \bar{N}  \sum_{i=1}^{N_{\text cell}} \sum_{i=1}^{N_{\text gal}}   \xi_{gv,\theta_\beta}(x_i,x_j) \frac{\Delta v_j}{\sigma^2_{v,j}} 
+ \sum_{i=1}^{N_{\text gal}}  \sum_{j=1}^{N_{\text gal}}   \frac{\Delta v_i}{\sigma^2_{v,i}}  \xi_{vv,\theta_\beta}(x_i,x_j)  \frac{\Delta v_j}{\sigma^2_{v,j}},\\
%\Delta \textbf{y}^T \textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \Delta \textbf{y} & \approx \sum_{i,j\in N_{gal}} N_i N_j \xi_{\delta\delta,\theta_\beta}(x_i,x_j) -2 \bar{N} \sum_{i\in N_{gal},j\in N_{cell}} \xi_{\delta\delta,\theta_\beta,ij}\\
%& +\bar{N}^2  \sum_{i,j\in N_{cell}}  \xi_{\delta\delta,\theta_\beta,ij} + 2 \sum_{i,j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta,ij} \\
%& -2\bar{N}\sum_{i\in N_{cell}, j\in N_{gal}} \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta,ij}  + \sum_{i,j\in N_{gal}} \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,\theta_\beta,ij} \frac{\Delta v_j}{\sigma_{v,j}^2}\\
%& = \sum_{i,j\in N_{gal}} \left( \xi_{\delta\delta,\theta_\beta}(x_i,x_j) + 2  \frac{\Delta v_j}{\sigma_{v,j}^2} \xi_{\delta v,\theta_\beta}(x_i,x_j) + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{vv,\theta_\beta}(x_i,x_j) \frac{\Delta v_j}{\sigma_{v,j}^2}\right)\\
%&-2  \sum_{i\in N_{gal}}\int n(x_i) \left(\xi_{\delta\delta,\theta_\beta}(x_i,y) + \frac{\Delta v_i}{\sigma_{v,i}^2} \xi_{v \delta,\theta_\beta}(x_i,y) \right) d^3y\\
%&  + \int n(x) n(y) \xi_{\delta\delta,\theta_\beta}(x,y)d^3xd^3y.
\end{align*}
\begin{align*}
\tr (\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{S}_\alpha) & 
= \bar{N}^2  \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text cell}} \xi_{gg,\theta_\beta}(x_i,x_j)\xi_{gg,\theta_\alpha}(x_j,x_i) 
+2 \bar{N}   \sum_{i=1}^{N_{\text cell}} \sum_{j=1}^{N_{\text gal}}\frac{1}{\sigma^{2}_{vj}}   \xi_{gv,\theta_\beta} (x_i,x_j)\xi_{\alpha,vg} (x_j,x_i)\\
& + \sum_{i=1}^{N_{\text gal}} \sum_{j=1}^{N_{\text gal}} \frac{1}{\sigma^{2}_{vi} \sigma^{2}_{vj}} \xi_{vv,\theta_\beta}(x_i,x_j)\xi_{vv,\theta_\alpha}(x_j,x_i) 
\end{align*}
\begin{align*}
\tr (\textbf{C}^{-1} \textbf{S}_\beta \textbf{C}^{-1} \textbf{C}_0) & 
=   \bar{N} \sum_{i=1}^{N_{\text cell}}  \xi_{gg,\theta_\beta}(x_i,x_i)  + \sum_{i=1}^{N_{\text gal}} \frac{1}{ \sigma^{2}_{vi}}\xi_{vv,\theta_\alpha}(x_i,x_i)
\end{align*}

The SN~Ia rate is $2.69 \times 10^{-5} \text{Mpc}^{-3} \text{yr}^{-1}$, meaning that after 10 years there
will be around 1 supernova every 3700~$\text{Mpc}^{3}$.
The Universe is $2.4 \times 10^9$~Mpc$^3$ out to $z=0.2$.  A rough bound on the number of necessary cells is $6\times 10^{5}$.
When taking the linear approximation the calculation scales as $N_{cell}^2$, so this is possible to do on a laptop.
The exact solution requires matrix factorizations that scale as $N_{cell}^3$, which is not so possible on a laptop.
